---
title: "Cell cycle prediction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load some packages

```{r load, warning=FALSE}

suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(MultiAssayExperiment))

suppressPackageStartupMessages(library(plotrix))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(gridExtra))

```

## Load Buettner data

From the Conquer database (http://imlspenticton.uzh.ch:3838/conquer/) the Buettner dataset was fetched:

1	EMTAB2805 (PMID 25599176)	Buettner2015	Mus musculus	mESC in different cell cycle stages	288	SMARTer C1	full-length

```{r data}
cc.data <- readRDS("/Users/erikh/Desktop/embryonic_data/EMTAB2805.rds")

# get tpms at gene level
exprs <- assays(experiments(cc.data)[["gene"]])[["TPM"]]
# remove the .1 etc from the ensembl gene name
gname <- rownames(exprs)
ensname <- gsub("\\.\\d+","",gname)
rownames(exprs) <- ensname

# remove spike-ins from expression table
ercc <- grep("ERCC",rownames(exprs))
exprs <- exprs[-ercc,]

# metadata with cell cycle stage assignment
M <- as.data.frame(colData(cc.data))
head(M)
```

## Filter out clearly low quality cells

Clearly some outlier cells on negative PC1 with low number of genes with pca using full dataset.

Mainly S-phase cells, even if most S-phase cells have a high median expression.

```{r filter}
# filter out cells with low nDet
nDet <- colSums(exprs>1)
hist(nDet,n=100)
# make a cutoff at 4000 genes, removes 16 cells
keep <- which(nDet>4000)

exprs<-exprs[,keep]
M <- M[keep,]

```



# Run cyclone


```{r cyclone, warning=F}
mmu.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))

set.seed(123)
sce <- SingleCellExperiment(assays=list(counts = exprs, logcounts = log2(exprs+1)))
cc <- cyclone(sce, pairs=mmu.pairs)

table(cc$phase)
table(M$cell_cycle_stage)


# plot overlap matrix:
source("/Users/erikh/Desktop/git/scrna-cell-cycle/r_scripts/cyclone_functions.R")
nS <- length(cc$phase)
o <- overlap_phyper2(split(1:nS,cc$phase),split(1:nS,M$cell_cycle_stage),plot=T)
```

Rows are the cyclone predictions, columns are the assigned phase from sorting.  

Predictions have slightly more G1 compared to their assignment, and not all cells are consistent.

## Plot onto PCA

```{r}
PC <- prcomp(t(log2(exprs+1)))
set.seed(123)

plot(data.frame(PC$x[,1:5]),col=as.factor(M$cell_cycle_stage),pch=16,main="Annotated phase")
plot(data.frame(PC$x[,1:5]),col=as.factor(cc$phase),pch=16,main="Cyclone prediction")


# also check if some PC is clearly driven by detected genes
nDet <- colSums(exprs>1)
colDet <- color.scale(nDet, c(0,1,1),c(1,1,0),0)
plot(data.frame(PC$x[,1:5]),col=colDet,pch=16)


```


PC1 is mainly due to gene detection, PC3 & 4 semms to separate the 3 phases best. 

# Cell cycle scoring in Seurat

Seurat only provides a list of cell cycle genes for human data (from Tirosh et al paper). First need to convert the human gene names to mouse orthologs and their ensembl gene ids.

```{r convert_names}
savefile = "seurat_cc_genes_mouse.Rdata"
if (file.exists(savefile)){
  load(savefile)
}else{

# Load human ensembl attributes
human = biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# Load mouse ensembl attributes
mouse = biomaRt::useMart("ensembl", dataset = "mmusculus_gene_ensembl")

hsa2mmEnsembl <- function(x,mouse,human){
  # Link both datasets and retrieve mouse genes from the human genes
  genes.list = biomaRt::getLDS(attributes = c("hgnc_symbol"), filters = "hgnc_symbol", values = x , mart = human, attributesL = c("mgi_symbol","ensembl_gene_id"), martL = mouse, uniqueRows = T)
  # Get unique names of genes 
  mouse.gene.list <- unique(genes.list[, 3])
  return(mouse.gene.list)
}

# load data from Seurat package
data(cc.genes)
# convert to mouse ensembl IDs
cc.genes.mouse <- lapply(cc.genes, hsa2mmEnsembl, mouse,human)
save(cc.genes.mouse, file=savefile)
}

```

Run Seurat CellCycleScoring function

```{r seurat}
# add cyclone prediction to metadata
M$cyclone <- cc$phases

# create a seurat object
sdata <- CreateSeuratObject(raw.data = exprs, min.cells = 3, min.genes = 200,
    project = "CC",is.expr=1,meta.data=M)

sdata <- NormalizeData(sdata)
sdata <- ScaleData(sdata)
sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5)

# run PCA
sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, do.print = FALSE)  
# predict cell cycle
sdata <- CellCycleScoring(sdata, g2m.genes = cc.genes.mouse$g2m.genes, 
                          s.genes = cc.genes.mouse$s.genes)

table(sdata@meta.data$Phase)

```

More G2M predictions with Seurat, less S-phase, similar for G1. 

```{r}
p1 <- PCAPlot(sdata, group.by = "cell_cycle_stage", plot.title="Annotation", do.return=T)
p2 <- PCAPlot(sdata, group.by = "Phase", plot.title="Seurat prediction", do.return=T)
p3 <- PCAPlot(sdata, group.by = "cyclone", plot.title="Cyclone prediction", do.return=T)

plot_grid(p1,p2,p3)
```

Clearly cyclone has better overlap with the annotation. But that data is used to train Seurat. 


Look at consistency in predictions per cell. 

```{r}

o <- overlap_phyper2(split(1:nS,cc$phase),split(1:nS,sdata@meta.data$Phase),plot=T)

# plot on pca
all <- as.matrix(sdata@meta.data[,c("Phase","cyclone")])
assigned <- apply(all,1,paste, collapse=":")
table(assigned)

sdata@meta.data$assigned <- assigned
PCAPlot(sdata, group.by = "assigned", plot.title="Combined")



```

Legend for PCA has Seurat prediction:Cyclone prediction. 


Clearly the S-phase assignment is the most inconsistent. Cells in S-phase with one method end up as G1 with the other. Also many S from cyclone are predicted as G2M with seurat. 

# PCA based on CC-genes

Take the gene sets from both cyclone and seurat and run pca with only those. See how the predictions look on that PCA.

```{r pca_cc}
cyclone.genes <- unique(unlist(lapply(mmu.pairs, unlist)))
length(cyclone.genes)
seurat.genes <- unlist(cc.genes.mouse)
length(seurat.genes)

# quite a few genes unique to seurat list
setdiff(seurat.genes,cyclone.genes)

genes.cc <- unique(c(cyclone.genes,seurat.genes))
length(genes.cc)

sdata.cc <- RunPCA(object = sdata, pc.genes = genes.cc, do.print = FALSE)

p1 <- PCAPlot(sdata.cc, group.by = "cell_cycle_stage", plot.title="Annotation",do.return=T)
p2 <- PCAPlot(sdata.cc, group.by = "Phase", plot.title="Seurat prediction", do.return=T)
p3 <- PCAPlot(sdata.cc, group.by = "cyclone", plot.title="Cyclone prediction", do.return=T)

plot_grid(p1,p2,p3)

```

Better separation of the G2M cells with this pca, but more mixing of G1/S cells. 

Also look at number of genes for the different groups. 

```{r}
p1 <- VlnPlot(sdata, features.plot = ("nGene"), group.by = "cell_cycle_stage", do.return =T)
p2 <- VlnPlot(sdata, features.plot = ("nGene"), group.by = "Phase", do.return=T)
p3 <- VlnPlot(sdata, features.plot = ("nGene"), group.by = "cyclone", do.return = T)

plot_grid(p1,p2,p3)
```

With the seurat predictions, less clear shift to low nDET for G1 cells. 

# Test on other datasets

From conquer fetched datasets:

5	GSE44183-GPL11154 (PMID 23892778)	Xue2013	Homo sapiens	embryonal development from oocyte to morula	29	Tang2010	full-length

6	GSE44183-GPL13112 (PMID 23892778)	Xue2013	Mus musculus	embryonal development from oocyte to morula	17	Tang2010	full-length	 

8 GSE45719 (PMID 24408435)	Deng2014	Mus musculus	development from zygote to blastocyst + adult liver	291	SMART-Seq	full-length

21	GSE63818-GPL16791 (PMID 26046443)	Guo2015	Homo sapiens	primordial germ cells from embryos at different times of gestation	328	Tang	full-length

24	GSE66507 (PMID 26293300)	Blakeley2015	Homo sapiens	preimplantation embryos	30	SMARTer	full-length

2 EMTAB3929 (PMID 27062923)	Petropoulos2016	Homo sapiens	preimplantation embryos from embryonic day 3 to 7	1529	Smart-Seq2	full-length

For each of these define some settings:

```{r}
datasets <- list()
datasets$XueHsa <- list(name = "GSE44183-GPL11154.rds", annot = "characteristics_ch1", species = "human")
datasets$XueMmu <- list(name = "GSE44183-GPL13112.rds", annot = "characteristics_ch1", species = "mouse")
datasets$GuoHsa <- list(name = "GSE63818-GPL16791.rds", annot = c("characteristics_ch1","source_name_ch1"), species = "human")
datasets$DengMmu <- list(name = "GSE45719.rds", annot = "source_name_ch1", species = "mouse")
datasets$BlakeleyHsa <- list(name = "GSE66507.rds", annot = "characteristics_ch1.1", species = "human")
datasets$PetropoulosHsa <- list(name = "EMTAB3929.rds", annot = "Characteristics.developmental.stage.", species = "human")
datasets$BuettnerMmu <- list(name = "EMTAB2805.rds", annot = "cell_cycle_stage", species = "mouse")

```

Translate seurat cc-genes to ensembl ids.

```{r}
savefile = "seurat_cc_genes_human.Rdata"
if (file.exists(savefile)){
  load(savefile)
}else{
  human = biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
  hsa2Ensembl <- function(x,human){
    genes.list = biomaRt::getBM(attributes = c("hgnc_symbol","ensembl_gene_id"), filters = "hgnc_symbol", values = x , mart = human, uniqueRows = T)
    # Get unique names of genes 
    ens.gene.list <- unique(genes.list[, 2])
    return(ens.gene.list)
  }
  # load data from Seurat package
  data(cc.genes)
  # convert to ensembl IDs
  cc.genes.human <- lapply(cc.genes, hsa2Ensembl, human)
  save(cc.genes.human, file=savefile)
}

```

Define a function for running all analysis on one dataset at a time.

```{r}
mmu.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))
hsa.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))


cc.analysis <- function(ds){
  print(ds)
  savefile <- paste("embryonic_data/processed/", ds, "_seurat_data.Rdata", sep="")
  if (!file.exists(savefile)){
  # load data
  file <- paste("embryonic_data/", datasets[[ds]]$name, sep='')
  cc.data <- readRDS(file)

  # get tpms at gene level
  exprs <- assays(experiments(cc.data)[["gene"]])[["TPM"]]
  # remove the .1 etc from the ensembl gene name
  gname <- rownames(exprs)
  ensname <- gsub("\\.\\d+", "", gname)
  rownames(exprs) <- ensname

  # remove spike-ins from expression table
  ercc <- grep("ERCC", rownames(exprs))
  exprs <- exprs[-ercc,]

  # metadata with cell cycle stage assignment
  M <- as.data.frame(colData(cc.data))
  head(M)  

  # run cyclone  
  set.seed(123)
  pairs <- hsa.pairs
  if (datasets[[ds]]$species == "mouse") { pairs <- mmu.pairs }
  sce <- SingleCellExperiment(assays=list(counts = exprs, logcounts = log2(exprs+1)))
  cc <- cyclone(sce, pairs=pairs)

  
  # create seurat object
  M$cyclone <- cc$phases
  sdata <- CreateSeuratObject(raw.data = exprs, min.cells = 3, min.genes = 200,
    project = "CC", is.expr=1, meta.data=M )
  sdata <- NormalizeData(sdata, display.progress = F)
  sdata <- ScaleData(sdata, display.progress = F)
  sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5, do.plot = F,
                           display.progress = F)
  pc.compute <- 20
  if (ncol(exprs) < 20){ pc.compute <- ncol(exprs)-1 }
  sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, 
                  do.print = FALSE, pcs.compute = pc.compute)  
  genes.use <- cc.genes.human
  if (datasets[[ds]]$species == "mouse") { genes.use <- cc.genes.mouse }
  sdata <- CellCycleScoring(sdata, g2m.genes = genes.use$g2m.genes, 
                          s.genes = genes.use$s.genes)
  save(sdata, file=savefile)
  }else{
    load(savefile)
  }
  print("cyclone:")
  print(table(sdata@meta.data$cyclone))
  print("Seurat:")
  print(table(sdata@meta.data$Phase))
  
  # plot onto pca
  annot <- datasets[[ds]]$annot
  p1 <- PCAPlot(sdata, group.by = "Phase", plot.title="Seurat prediction", do.return=T)
  p2 <- PCAPlot(sdata, group.by = "cyclone", plot.title="Cyclone prediction", do.return=T)
  p3 <- PCAPlot(sdata, group.by = annot[1], plot.title="Annotation", do.return=T) +
     theme(legend.text=element_text(size=4))
  if (length(annot) == 2){
    p4 <- PCAPlot(sdata, group.by = annot[2], plot.title="Annotation", do.return=T) +
     theme(legend.text=element_text(size=4))
    pl <- plot_grid(p1,p2,p3,p4)
  }else {
    pl <- plot_grid(p1,p2,p3)
  }
  print(pl)
}

```


Run for each of the datasets

```{r run_all}
for (ds in names(datasets)){
  cc.analysis(ds)
}
```

* XueHsa - More S-phase with Seurat, Only G1 in the later stages (4-cell to morula). Cyclone precits mainly G1, some G2M for later stages. 
* XueMmu - Seurat - G1 & G2M (G1 at later stages), Cyclone S + G2M (Mainly S at middle stages)
* GuoHsa - Seurat - mainly G1, Cyclone - mainly S. Somatic cells mainly G1 with Seurat, mainly S with cyclone. P germ cells - more S at later stages with Seurat, mixed with cyclone. 
* DengMmu - Seurat - mainly G1, Cyclone mainly S. Seurat - all early stages G2M, 8-16 mainly G1, blastocyst mixed. Cyclone - early stages mixed G2M/S, 8-16 S+G1, blastocyst mixed. 
* BlakeleyHsa - Seurat mixed, Cyclon mainly G1. TE mainly G1 with cycone, S+G2M with seurat. 
* PetropoulosHsa - mainly G1 with both, but more G2M/S with Seurat. Earlier stages  G2M with seurat. Negative PC2 mainly S/G2M with cyclone, Positive PC2 mainly S/G2M with seurat. Clearly opposite pattern. 

### Check cell paper data

Since the CP data is a clear example where cyclone and seurat disagrees. Check some marker genes. 

```{r}
ds <- "PetropoulosHsa"
savefile <- paste("embryonic_data/processed/", ds, "_seurat_data.Rdata", sep="")
load(savefile)

# AURKA:ENSG00000087586, PLK1:ENSG00000166851 - G2 marker
# KIF18B:ENSG00000186185, AURKB:ENSG00000178999 S+G2M
# CCND3;ENSG00000112576, CCNE2:ENSG00000175305  - G1-S

FeaturePlot(sdata, features.plot = c("ENSG00000087586","ENSG00000166851", "ENSG00000186185","ENSG00000178999", "ENSG00000112576", "ENSG00000175305"), reduction.use = "pca")
     
# PGF:ENSG00000119630, SERPINB9:ENSG00000170542 - mural TE marker
# VCAN:ENSG00000038427, IL6R:ENSG00000160712 - ICM markers                                    
FeaturePlot(sdata, features.plot = c( "nGene","ENSG00000119630", "ENSG00000170542", "ENSG00000038427", "ENSG00000160712"), reduction.use = "pca")



```

Looks like the cells in upper part have more of the G2M/S markers - perhaps seurat is correct in assignment? Should be short/no G1 at the early stages until differentiation begins, while at specification of lineages the G1 phase becomes longer.  Not sure at what embryonic stage in human this shift occurs. 

#### session info
```{r session}
sessionInfo()
```