---
title: "protocol"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Protocol for evaluation of different cell cycle programs on different datasets.

The script is divided into different parts. In each part a different dataset is loaded and used as input for the programs; Cyclone, Seurat, reCAT and IA-SVA. The last part of the script is an evaluation of the performance of the programs. After the datasets that we have included here there is a section where the user can insert their own dataset and run it. In order to include the custom data into the evaluation it needs to be annotated according to cell cycle beforehand. See instructions above the section to see which format needs to be used.

### Loading packages and marker genes (for Seurat)

```{r load, warning=FALSE}

suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(MultiAssayExperiment))

suppressPackageStartupMessages(library(plotrix))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(gridExtra))

```

Run this chunk for Seurat to find the marker genes for mouse and human. For other organisms a Sandbag function needs to be used. It will be saved to the environment. If the variables were to be cleant this chunk needs to be run again. 
```{r convert_names}
savefile = "seurat_cc_genes_mouse.Rdata"
if (file.exists(savefile)){
  load(savefile)
}else{

# Load human ensembl attributes
human = biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# Load mouse ensembl attributes
mouse = biomaRt::useMart("ensembl", dataset = "mmusculus_gene_ensembl")

hsa2mmEnsembl <- function(x,mouse,human){
  # Link both datasets and retrieve mouse genes from the human genes
  genes.list = biomaRt::getLDS(attributes = c("hgnc_symbol"), filters = "hgnc_symbol", values = x , mart = human, attributesL = c("mgi_symbol","ensembl_gene_id"), martL = mouse, uniqueRows = T)
  # Get unique names of genes 
  mouse.gene.list <- unique(genes.list[, 3])
  return(mouse.gene.list)
}

# load data from Seurat package
data(cc.genes)
# convert to mouse ensembl IDs
cc.genes.mouse <- lapply(cc.genes, hsa2mmEnsembl, mouse,human)
save(cc.genes.mouse, file=savefile)
}

```


## Buttner *et al.* data (2015)
Section for running the Buttner data through the programs.

### Load the Buttner data
The Buttner data can be downloaded from the the conquer data base (http://imlspenticton.uzh.ch:3838/conquer/)

```{r Buttner data}
cc.data <- readRDS("/Users/Lovisa Lundeberg/Documents/scrna-cell-cycle/r_scripts/lovisa/embryonic_data/EMTAB2805.rds")

# get tpms at gene level
exprs <- assays(experiments(cc.data)[["gene"]])[["TPM"]]
# remove the .1 etc from the ensembl gene name
gname <- rownames(exprs)
ensname <- gsub("\\.\\d+","",gname)
rownames(exprs) <- ensname

# remove spike-ins from expression table
ercc <- grep("ERCC",rownames(exprs))
exprs <- exprs[-ercc,]

# metadata with cell cycle stage assignment
M <- as.data.frame(colData(cc.data))
```

### Filter out cells of low quality


```{r filter}
# filter out cells with low nDet
nDet <- colSums(exprs>1)
hist(nDet,n=100)
keep <- which(nDet>4000)

exprs<-exprs[,keep]
M <- M[keep,]

```

### Running cyclone

```{r cyclone, warning=F}
mmu.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))

set.seed(123)
sce <- SingleCellExperiment(assays=list(counts = exprs, logcounts = log2(exprs+1)))
cc <- cyclone(sce, pairs=mmu.pairs)

table(cc$phase)

# saving the result as a numeric vector
cyc_bu_pre <- as.vector(unlist(lapply(c(table(cc$phases)), as.numeric)))
```

### Running Seurat

```{r seurat}
# add cyclone prediction to metadata
M$cyclone <- cc$phases

# create a seurat object
sdata <- CreateSeuratObject(raw.data = exprs, min.cells = 3, min.genes = 200,
    project = "CC",is.expr=1,meta.data=M)

sdata <- NormalizeData(sdata)
sdata <- ScaleData(sdata)
sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5)

# run PCA
sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, do.print = FALSE)  
# predict cell cycle
sdata <- CellCycleScoring(sdata, g2m.genes = cc.genes.mouse$g2m.genes, 
                          s.genes = cc.genes.mouse$s.genes)

table(sdata@meta.data$Phase)

# saving the result as a numeric vector
seu_bu_pre <- as.vector(unlist(lapply(c(table(sdata@meta.data$Phase)), as.numeric)))
```

### Running reCAT

# Clears out all genes not present as marker genes

```{r reCAT_preprocess}

# Log transforms the data as get_test requires log-transformed data 
exprs <- log(exprs+1)

source("reCAT_master/R/get_test_exp.R")
exprs <- get_test_exp(exprs)
dim(exprs)
```

# Orders the data using the TSP algorithm

```{r reCAT_order}
source("reCAT_master/R/get_ordIndex.R")
# exprs = input matrix, 3 = no. threads used for processing, can be increased on cluster
nthread =3

ordIndex <- get_ordIndex(exprs, nthread)
```

# Get cell-cycle scores using the Bayes-scores & Mean-scores algorithm

```{r reCAT_scores}
source("reCAT_master/R/get_score.R")
score_result <- get_score(t(exprs))
score_result$bayes_score
score_result$mean_score
# cat(score_result$bayes_score, file="bayes_score.txt")
# cat(score_result$mean_score, file="mean_score.txt")
```

# Plots the score values 

```{r reCAT_plot_scores}
source("reCAT_master/R/plot.R")
plot_bayes(score_result$bayes_score, ordIndex)
plot_mean(score_result$mean_score, ordIndex)
```


```{r reCAT_rdata}
source("reCAT_master/R/get_rdata.R")
rdata <- get_rdata(score_result, ordIndex)
View(rdata)
```

# Finds the start of the cell-cycle for time-series scoring

```{r reCAT_get_start}
source("reCAT_master/R/get_start.R")
cycle_start = get_start(score_result, ordIndex)
print(cycle_start)
```

```{r reCAT_get_myord}
source("reCAT_master/R/get_myord.R")
myord = get_myord(cycle_start, ordIndex)
```

# Finds the cell cycle phases

```{r reCAT_find_phase}
source("reCAT_master/R/get_hmm.R")
cls_num <- 3
hmm_result <- get_hmm_order(bayes_score = score_result$bayes_score, mean_score = score_result$mean_score, ordIndex = ordIndex, cls_num = cls_num, myord = myord, rdata = rdata)

g1_c = sum(hmm_result == 1)
s_c = sum(hmm_result == 2)
g2m_c = sum(hmm_result == 3)
phase_counts = t(data.frame(g1_c, g2m_c, s_c))
write.table(phase_counts,"recat_phase_output.txt",sep="\t",col.names = FALSE, row.names=FALSE)
View(phase_counts)
```

## Leng *et al* data (2015)
Section for running the Leng *et al* data through the programs.

### Load the Leng *et al* data
The dataset consists of 247 cells with cell cycle annotation 91 cells in G1, 80 in S and 76 in G2M.

The data can be downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE64016

```{r data}
#reading in datafile
exprs_le <- read.csv("/Users/Lovisa Lundeberg/Documents/scrna-cell-cycle/r_scripts/lovisa/embryonic_data/GSE64016_H1andFUCCI_normalized_EC.csv", header = TRUE, sep = ",", quote = "\"")

#saving the columns that are labeled (215-461) into new matrix
exprs_le <- exprs_le[,c(1,215:461)]

#adding rownames that are in the first column of the original data
rownames(exprs_le) <- paste(exprs_le[,c(1)])
exprs_le <- exprs_le[,-1]
```

Translating the HGNC symbols that the data have originally to ensembl IDs and adding it as a column in the matrix. This is needed in order to run Cyclone.
```{r database}
# getting ensembl ids based on the hgnc symbols in the file
library(biomaRt)
mart = useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
ens_id_hum = getBM(filters = "hgnc_symbol", attributes =       c("ensembl_gene_id","hgnc_symbol"), values=rownames(exprs_le), mart = mart)

# removing all genes that lack ensembl id or have multiple ones
ensCount =  table(ens_id_hum$ensembl_gene_id) 
one2one = names(which(ensCount == 1))
ensCount =  table(ens_id_hum$hgnc_symbol) 
one2oneGene = names(which(ensCount == 1))

ens_id_hum =  ens_id_hum[ens_id_hum$ensembl_gene_id %in% one2one, ]
ens_id_hum =  ens_id_hum[ens_id_hum$hgnc_symbol %in% one2oneGene, ]
exprs_le = exprs_le[rownames(exprs_le) %in% ens_id_hum$hgnc_symbol, ]
exprs_le = exprs_le[ens_id_hum$hgnc_symbol, ]


#adding ensembl IDs as the first column in the data, removing NAs in the data
exprs_le <- cbind(ensembl_id=ens_id_hum$ensembl_gene_id,exprs_le)
exprs_le <- na.omit(exprs_le)
```

### Filtering out cells of low quality

```{r filter}
# filter out cells with low nDet
nDet <- colSums(exprs_le[,-1]>1)
hist(nDet,n=100)
keep <- which(nDet>4000)

exprs_le<-exprs_le[,keep]
```

### Running cyclone

```{r cyclone, warning=F}
#adding ensembl ids as rownames
exprs_id <- exprs_le[,-1]
rownames(exprs_id) <- paste(exprs_le[,1]) 

exprs_id <- data.matrix(exprs_id)
mmu.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))

# running cyclone
set.seed(123)
sce_leng <- SingleCellExperiment(assays=list(counts = exprs_id, logcounts = log2(exprs_id+1)))
cc <- cyclone(sce_leng, pairs=mmu.pairs)

table(cc$phase)

# saving the result in a numeric vector
cyc_le_pre <- as.vector(unlist(lapply(c(table(cc$phases)), as.numeric)))
```

### Running Seurat

```{r seurat}
#editing the matrix to have hgnc symbols
exprs_hgnc <- exprs_le[,-1]

sdata <- CreateSeuratObject(raw.data = exprs_hgnc, min.cells = 3, min.genes = 200,
    project = "testseurat",is.expr=1,meta.data=NULL)

sdata <- NormalizeData(sdata)
sdata <- ScaleData(sdata)
sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5)


# run PCA
sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, do.print = FALSE)  
# predict cell cycle
sdata <- CellCycleScoring(sdata, g2m.genes = cc.genes$g2m.genes, 
                          s.genes = cc.genes$s.genes)

table(sdata@meta.data$Phase)
seu_le_pre <- as.vector(unlist(lapply(c(table(sdata@meta.data$Phase)), as.numeric)))
```

### Running reCAT

# Clears out all genes not present as marker genes

```{r reCAT_preprocess}

# Log transforms the data as get_test requires log-transformed data 
exprs_le <- log(exprs_le+1)

source("reCAT_master/R/get_test_exp.R")
exprs_le <- get_test_exp(exprs_le)
dim(exprs_le)
```

# Orders the data using the TSP algorithm

```{r reCAT_order}
source("reCAT_master/R/get_ordIndex.R")
# exprs_le = input matrix, 3 = no. threads used for processing, can be increased on cluster
nthread =3

ordIndex <- get_ordIndex(exprs_le, nthread)
```

# Get cell-cycle scores using the Bayes-scores & Mean-scores algorithm

```{r reCAT_scores}
source("reCAT_master/R/get_score.R")
score_result <- get_score(t(exprs_le))
score_result$bayes_score
score_result$mean_score
# cat(score_result$bayes_score, file="bayes_score.txt")
# cat(score_result$mean_score, file="mean_score.txt")
```

# Plots the score values 

```{r reCAT_plot_scores}
source("reCAT_master/R/plot.R")
plot_bayes(score_result$bayes_score, ordIndex)
plot_mean(score_result$mean_score, ordIndex)
```


```{r reCAT_rdata}
source("reCAT_master/R/get_rdata.R")
rdata <- get_rdata(score_result, ordIndex)
View(rdata)
```

# Finds the start of the cell-cycle for time-series scoring

```{r reCAT_get_start}
source("reCAT_master/R/get_start.R")
cycle_start = get_start(score_result, ordIndex)
print(cycle_start)
```

```{r reCAT_get_myord}
source("reCAT_master/R/get_myord.R")
myord = get_myord(cycle_start, ordIndex)
```

# Finds the cell cycle phases

```{r reCAT_find_phase}
source("reCAT_master/R/get_hmm.R")
cls_num <- 3
hmm_result <- get_hmm_order(bayes_score = score_result$bayes_score, mean_score = score_result$mean_score, ordIndex = ordIndex, cls_num = cls_num, myord = myord, rdata = rdata)

g1_c = sum(hmm_result == 1)
s_c = sum(hmm_result == 2)
g2m_c = sum(hmm_result == 3)
phase_counts = t(data.frame(g1_c, g2m_c, s_c))
write.table(phase_counts,"recat_phase_output.txt",sep="\t",col.names = FALSE, row.names=FALSE)
View(phase_counts)
```

## Sasagawa *et al* data (2013)
Section for running the data from Sasagawa *et al*.

### Load the Sasagawa *et al* data
The data is a small dataset of 23 cells. The ES files with a clear cell annotation is used (G1, S and M). The annotation is 8 cells in G1, 7 in S and 8 in G2M.

The data can be downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE42268. This is 23 txt files. In this code they are all put into a folder called "sasagawa" where all the txt files are loaded from.

```{r data}
mypath = "/Users/Lovisa Lundeberg/Documents/scrna-cell-cycle/r_scripts/lovisa/embryonic_data/sasagawa"
setwd(mypath)

# reading in the separate text files and putting them next to each other in a data frame
txt_files_ls = list.files(path=mypath, pattern="*.txt")
txt_files_df <- lapply(txt_files_ls, function(x) {read.table(file = x, header = T, sep ="")})
exprs_sa <- do.call("cbind", lapply(txt_files_df, as.data.frame))

# saving the ids
id <- as.data.frame(exprs_sa[,1:2])

# removing the rows with ids from the matrix
exprs_sa <- exprs_sa[ ,-which(names(exprs_sa) %in% c("id", "gene.symbol"))]

# adding the ids back into the matrix
rownames(exprs_sa) <- paste(id[,1]) 
exprs_sa <- cbind(hgnc_symbol=id[,2],exprs_sa)
```

### Filtering out cells of low quality

```{r filter}
# filter out cells with low nDet
nDet <- colSums(exprs_sa[,-1]>1)
hist(nDet,n=100)
keep <- which(nDet>4000)

exprs_sa <- exprs_sa[,keep]
```

### Running Cyclone

```{r cyclone, warning=F}
# editing the matrix to have only ensembl ids
exprs_sa_id <- data.matrix(exprs_sa[,-1])
mmu.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))

# running cyclone
set.seed(123)
sce_leng <- SingleCellExperiment(assays=list(counts = exprs_sa_id, logcounts = log2(exprs_sa_id+1)))
cc <- cyclone(sce_leng, pairs=mmu.pairs)

table(cc$phase)

# saving the result as a numeric vector
cyc_sa_pre <- as.vector(unlist(lapply(c(table(cc$phases)), as.numeric)))
```

### Running Seurat

```{r Seurat}
# create a seurat object
sdata <- CreateSeuratObject(raw.data = exprs_sa_id, min.cells = 3, min.genes = 200,
    project = "CC",is.expr=1,meta.data=NULL)

sdata <- NormalizeData(sdata)
sdata <- ScaleData(sdata)
sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5)

# run PCA
sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, do.print = FALSE)  
# predict cell cycle
sdata <- CellCycleScoring(sdata, g2m.genes = cc.genes.mouse$g2m.genes, 
                          s.genes = cc.genes.mouse$s.genes)

table(sdata@meta.data$Phase)

# saving the result as a numeric vector
seu_sa_pre <- as.vector(unlist(lapply(table(sdata@meta.data$Phase), as.numeric)))
```

## Evaluation
This section is for evaluating the result of the programs. In order to run this section all programs have to have been run using all the datasets.

### RMSE
This section makes plots for each of the programs of the RMSE values. It is calculated with the result vectors and the annotation of the cells.

```{r data}
# creating vectors of the annotations
bu_an <- c(95, 96, 81)
le_an <- c(91, 76, 80)
sa_an <- c(7, 8, 7)
```

Calculatin the RMSE value for the programs using all of the datasets.
```{r RMSE}
# function for calculating RMSE
RMSE = function(m, o){
  sqrt(mean((m - o)^2))
}

# calculating RMSE for cyclone
RMSE_cyc <- c(RMSE(cyc_bu_pre, bu_an), RMSE(cyc_le_pre, le_an), RMSE(cyc_sa_pre, sa_an))
 
# calculating RMSE for Seurat
RMSE_seu <- c(RMSE(seu_bu_pre, bu_an), RMSE(seu_le_pre, le_an), RMSE(seu_sa_pre, sa_an))
```

Making barplots of the RMSE values.
```{r plot}
barplot(RMSE_seu, main = "Seurat", xlab ="Data", ylab="RMSE", names.arg=c("Buttner et al.","Leng et al.", "Sasagawa et al."), border="blue", density=c(10), col=c("purple","purple","purple"))


barplot(RMSE_cyc, main = "Cyclone", xlab ="Data", ylab="RMSE", names.arg=c("Buttner et al.","Leng et al.", "Sasagawa et al."), border="blue", density=c(10), col=c("purple","purple","purple"))
```

### Running reCAT

# Clears out all genes not present as marker genes

```{r reCAT_preprocess}

# Log transforms the data as get_test requires log-transformed data 
exprs_sa <- log(exprs_sa+1)

source("reCAT_master/R/get_test_exp.R")
exprs_sa <- get_test_exp(exprs_sa)
dim(exprs_sa)
```

# Orders the data using the TSP algorithm

```{r reCAT_order}
source("reCAT_master/R/get_ordIndex.R")
# exprs_sa = input matrix, 3 = no. threads used for processing, can be increased on cluster
nthread =3

ordIndex <- get_ordIndex(exprs_sa, nthread)
```

# Get cell-cycle scores using the Bayes-scores & Mean-scores algorithm

```{r reCAT_scores}
source("reCAT_master/R/get_score.R")
score_result <- get_score(t(exprs_sa))
score_result$bayes_score
score_result$mean_score
# cat(score_result$bayes_score, file="bayes_score.txt")
# cat(score_result$mean_score, file="mean_score.txt")
```

# Plots the score values 

```{r reCAT_plot_scores}
source("reCAT_master/R/plot.R")
plot_bayes(score_result$bayes_score, ordIndex)
plot_mean(score_result$mean_score, ordIndex)
```


```{r reCAT_rdata}
source("reCAT_master/R/get_rdata.R")
rdata <- get_rdata(score_result, ordIndex)
View(rdata)
```

# Finds the start of the cell-cycle for time-series scoring

```{r reCAT_get_start}
source("reCAT_master/R/get_start.R")
cycle_start = get_start(score_result, ordIndex)
print(cycle_start)
```

```{r reCAT_get_myord}
source("reCAT_master/R/get_myord.R")
myord = get_myord(cycle_start, ordIndex)
```

# Finds the cell cycle phases

```{r reCAT_find_phase}
source("reCAT_master/R/get_hmm.R")
cls_num <- 3
hmm_result <- get_hmm_order(bayes_score = score_result$bayes_score, mean_score = score_result$mean_score, ordIndex = ordIndex, cls_num = cls_num, myord = myord, rdata = rdata)

g1_c = sum(hmm_result == 1)
s_c = sum(hmm_result == 2)
g2m_c = sum(hmm_result == 3)
phase_counts = t(data.frame(g1_c, g2m_c, s_c))
write.table(phase_counts,"recat_phase_output.txt",sep="\t",col.names = FALSE, row.names=FALSE)
View(phase_counts)
```