---
title: "protocol"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Protocol for evaluation of different cell cycle programs on different datasets.

The script is divided into different parts. In each part a different dataset is loaded and used as input for the programs; Cyclone, Seurat and reCAT. The last part of the script is an evaluation of the performance of the programs. 

A new section with other data can be included in a similar fashion. However, this new data needs to be annotated according to cell cycle beforehand in order for the evaluation to work. 

See details above the section to see which format needs to be used. Before starting to run the script a working directory needs to be set by using setwd("path") into a folder containing two folders named "embryonic_data" and "reCAT_master". The embryonic_data folder is where all the data should be put and the reCAT_master should contain all the scripts that reCAT needs in order to run. These folders should come automatically when cloning the GitHub. The data files needs to be downloaded separately and put into the folder.


References
Cyclone was developed by MarioniLab/Scran 2015. 
https://github.com/MarioniLab/scran/
Seurat was by Saijita lab 2015.
https://github.com/satijalab/seurat
reCAT was developed by Tinglab 2017, modified by Hugo Swenson 2018. 
https://github.com/tinglab/reCAT

### Loading packages and marker genes (for Seurat)

```{r load, warning=FALSE}

suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(plotrix))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(gridExtra))

```

Run this chunk for Seurat to find the marker genes for mouse and human. It will be saved to the environment. 

This includes genes for human and mouse.
```{r convert names}
savefile <- "seurat_cc_genes_mouse.Rdata"
if (file.exists(savefile)){
  load(savefile)
}else{

# load human ensembl attributes
human = biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# load mouse ensembl attributes
mouse = biomaRt::useMart("ensembl", dataset = "mmusculus_gene_ensembl")

hsa2mmEnsembl <- function(x,mouse,human){
  # link both datasets and retrieve mouse genes from the human genes
  genes.list = biomaRt::getLDS(attributes = c("hgnc_symbol"), filters = "hgnc_symbol", values = x , mart = human, attributesL = c("mgi_symbol","ensembl_gene_id"), martL = mouse, uniqueRows = T)
  # get unique names of genes 
  mouse.gene.list <- unique(genes.list[, 3])
  return(mouse.gene.list)
}

# load data from Seurat package
data(cc.genes)
# convert to mouse ensembl IDs
cc.genes.mouse <- lapply(cc.genes, hsa2mmEnsembl, mouse,human)
save(cc.genes.mouse, file=savefile)
}

```


## Buettner *et al.* data (2015)
Section for running the Buettner data through the programs.

### Load the Buettner data
The Buettner data can be downloaded from the the conquer data base (http://imlspenticton.uzh.ch:3838/conquer/).

This file is in rds format and no additional gene IDs than the ones provided in the file are needed. 

After filtration, if done as suggested below, the Buettner data contains 272 cells with the annotation 95 cells in G1, 81 in S and 95 in G2M.

```{r Buettner data}
path <- getwd()
path_bu <- paste(path, "/embryonic_data/buettner/EMTAB2805.rds", sep = "")

cc.data <- readRDS(path_bu)

# get tpms at gene level
exprs <- assays(experiments(cc.data)[["gene"]])[["TPM"]]
# remove the .1 etc from the ensembl gene name
gname <- rownames(exprs)
ensname <- gsub("\\.\\d+","",gname)
rownames(exprs) <- ensname

# remove spike-ins from expression table
ercc <- grep("ERCC",rownames(exprs))
exprs <- exprs[-ercc,]

# metadata with cell cycle stage assignment
M <- as.data.frame(colData(cc.data))
```

### Filter out cells of low quality
See report for details about the filtration step.
```{r filter Buettner}
# filter out cells with low nDet
nDet <- colSums(exprs>1)
hist(nDet,n=100)
keep <- which(nDet>4000)

exprs<-exprs[,keep]
M <- M[keep,]

```

### Running cyclone
Cyclone requires a matrix with ensembl IDs as rownames for mice data.
```{r cyclone Buettner}
mmu.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))

set.seed(123)
sce <- SingleCellExperiment(assays=list(counts = exprs, logcounts = log2(exprs+1)))
cc <- cyclone(sce, pairs=mmu.pairs)

table(cc$phase)


# saving the result as a numeric vector
cyc_bu_pre <- as.vector(unlist(lapply(c(table(cc$phases)), as.numeric)))
bu_cyc_vec <- unlist(cc$phases)
```

### Running Seurat
Seurat requires ensembl IDs for mice data.
```{r seurat Buettner}
# add cyclone prediction to metadata
M$cyclone <- cc$phases

# create a seurat object
sdata <- CreateSeuratObject(raw.data = exprs, min.cells = 3, min.genes = 200,
    project = "CC",is.expr=1,meta.data=M)

sdata <- NormalizeData(sdata)
sdata <- ScaleData(sdata)
sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5)

# run PCA
sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, do.print = FALSE)  
# predict cell cycle
sdata <- CellCycleScoring(sdata, g2m.genes = cc.genes.mouse$g2m.genes, 
                          s.genes = cc.genes.mouse$s.genes)

table(sdata@meta.data$Phase)

# saving the result as a vectors
seu_bu_pre <- as.vector(unlist(lapply(c(table(sdata@meta.data$Phase)), as.numeric)))
bu_seu_vec <- unlist(sdata@meta.data$Phase)
```

### Running reCAT
reCAT requires a matrix and ensembl IDs for mice data.

Clears out all genes not present as marker genes
```{r preprocess Buettner}
# log transforms the data as get_test requires log-transformed data 
exprs_log_bu <- log(exprs+1)

source("reCAT_master/R/get_test_exp.R")
exprs_log_bu <- get_test_exp(exprs_log_bu)
```

Orders the data using the TSP algorithm
```{r order Buettner}
source("reCAT_master/R/get_ordIndex.R")
# exprs = input matrix, 3 = no. threads used for processing, can be increased on cluster
nthread =3
ordIndex_bu <- get_ordIndex(exprs_log_bu, nthread)
```

Get cell-cycle scores using the Bayes-scores & Mean-scores algorithm
```{r scores Buettner}
source("reCAT_master/R/get_score.R")
score_result_bu <- get_score(t(exprs_log_bu))

# plots the score values
source("reCAT_master/R/plot.R")
plot_bayes(score_result_bu$bayes_score, ordIndex_bu)
```

```{r rdata Buettner}
# finds the start of the cell-cycle for time-series scoring
source("reCAT_master/R/get_start.R")
cycle_start_bu = get_start(score_result_bu, ordIndex_bu)
print(cycle_start_bu)

# find best region for assessing each phase
source("reCAT_master/R/get_rdata.R")
rdata_bu <- get_rdata(score_result_bu, ordIndex_bu)
print(rdata_bu)

# obtain the real time cycle
source("reCAT_master/R/get_myord.R")
myord_bu = get_myord(cycle_start_bu, ordIndex_bu)

```

Finds the cell cycle phases
```{r classification Buettner}
source("reCAT_master/R/get_hmm.R")
cls_num <- 3
hmm_result_bu <- get_hmm_order(bayes_score = score_result_bu$bayes_score, mean_score = score_result_bu$mean_score, ordIndex = ordIndex_bu, cls_num = cls_num, myord = myord_bu, rdata = rdata_bu)

g1_c = sum(hmm_result_bu == 1)
s_c = sum(hmm_result_bu == 2)
g2m_c = sum(hmm_result_bu == 3)


# saving results into vectors
recat_bu_pre <- c(g1_c,g2m_c,s_c)

# translating from numbers into letters
data_comp <- data.frame(hmm_result_bu[order(ordIndex_bu)], rownames(exprs_log_bu))
bu_rec_vec <- gsub("1", "G1", data_comp[[1]])
bu_rec_vec <- gsub("2", "S", bu_rec_vec)
bu_rec_vec <- gsub("3", "G2M", bu_rec_vec)

table(bu_rec_vec)
```

## Leng *et al* data (2015)
Section for running the Leng data through the programs.

### Load the Leng data
The dataset consists of 247 human cells with cell cycle annotation 91 cells in G1, 80 in S and 76 in G2M. The filtration step is present for this data as well so it can be changed if desired even though no cells are removed when running it as it is now.

The data can be downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE64016

```{r data Leng}
# reading in datafile
path_le <- paste(path, "/embryonic_data/leng/GSE64016_H1andFUCCI_normalized_EC.csv", sep = "")

exprs_le <- read.csv(path_le, header = TRUE, sep = ",", quote = "\"")

# saving the columns that are labelled (215-461) into new matrix
exprs_le <- exprs_le[,c(1,215:461)]

# adding rownames that are in the first column of the original data
rownames(exprs_le) <- paste(exprs_le[,c(1)])
exprs_le <- exprs_le[,-1]
```

Translating the HGNC symbols that the data have originally to ensembl IDs and adding it as a column in the matrix. This is needed in order to run Seurat.
```{r database}
# getting ensembl ids based on the hgnc symbols in the file
library(biomaRt)
mart = useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
ens_id_hum = getBM(filters = "hgnc_symbol", attributes =       c("ensembl_gene_id","hgnc_symbol"), values=rownames(exprs_le), mart = mart)

# removing all genes that lack ensembl id or have multiple ones
ensCount =  table(ens_id_hum$ensembl_gene_id) 
one2one = names(which(ensCount == 1))
ensCount =  table(ens_id_hum$hgnc_symbol) 
one2oneGene = names(which(ensCount == 1))

ens_id_hum =  ens_id_hum[ens_id_hum$ensembl_gene_id %in% one2one, ]
ens_id_hum =  ens_id_hum[ens_id_hum$hgnc_symbol %in% one2oneGene, ]
exprs_le = exprs_le[rownames(exprs_le) %in% ens_id_hum$hgnc_symbol, ]
exprs_le = exprs_le[ens_id_hum$hgnc_symbol, ]

# adding ensembl ids
exprs_le <- cbind(ensembl_id=ens_id_hum$ensembl_gene_id,exprs_le)

# removing NAs in the data
exprs_le <- na.omit(exprs_le)

```
### Filtering out cells of low quality
See report for details.

```{r filter Leng}
# filter out cells with low nDet
id <- exprs_le[,1]
exprs_le <- exprs_le[,-1]
nDet <- colSums(exprs_le>1)
hist(nDet,n=100)
keep <- which(nDet>4000)

exprs_le<-exprs_le[,keep]

exprs_le <-cbind(ensembl_id=id, exprs_le)
```

### Running cyclone
Cyclone requires a matrix with ensembl IDs as input for human data.
```{r cyclone Leng}
#adding ensembl ids as rownames
exprs_le_id <- exprs_le[,-1]
rownames(exprs_le_id) <- paste(exprs_le[,1]) 

exprs_le_id <- data.matrix(exprs_le_id)
mmu.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))

# running cyclone
set.seed(123)
sce <- SingleCellExperiment(assays=list(counts = exprs_le_id, logcounts = log2(exprs_le_id+1)))
cc <- cyclone(sce, pairs=mmu.pairs)

table(cc$phase)


# saving the results as vectors
cyc_le_pre <- as.vector(unlist(lapply(c(table(cc$phases)), as.numeric)))
le_cyc_vec <- unlist(cc$phases)
```

### Running Seurat
Seurat requires HGNC symbols for human data.
```{r seurat Leng}
# editing the matrix to have hgnc symbols
exprs_hgnc <- exprs_le[,-1]

sdata <- CreateSeuratObject(raw.data = exprs_hgnc, min.cells = 3, min.genes = 200,
    project = "testseurat",is.expr=1,meta.data=NULL)

sdata <- NormalizeData(sdata)
sdata <- ScaleData(sdata)
sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5)


# run PCA
sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, do.print = FALSE)  
# predict cell cycle
sdata <- CellCycleScoring(sdata, g2m.genes = cc.genes$g2m.genes, 
                          s.genes = cc.genes$s.genes)

table(sdata@meta.data$Phase)

# saving the results as vectors
seu_le_pre <- as.vector(unlist(lapply(c(table(sdata@meta.data$Phase)), as.numeric)))
le_seu_vec <- unlist(sdata@meta.data$Phase)
```

### Running reCAT
reCAT requires a matrix with ensembl IDs for human data.

Clears out all genes not present as marker genes
```{r preprocess Leng}
# log transforms the data as get_test requires log-transformed data 
exprs_le_log <- log(exprs_le_id+1)

source("reCAT_master/R/get_test_exp.R")
exprs_le_log <- get_test_exp(exprs_le_log)
```

Orders the data using the TSP algorithm

```{r order Leng}
source("reCAT_master/R/get_ordIndex.R")
# exprs_le_log = input matrix, 3 = no. threads used for processing, can be increased on cluster
nthread =3
ordIndex_le <- get_ordIndex(exprs_le_log, nthread)
```

Get cell-cycle scores using the Bayes-scores & Mean-scores algorithm
```{r scores Leng}
source("reCAT_master/R/get_score.R")
score_result_le <- get_score(t(exprs_le_log))

# plots the score values 
source("reCAT_master/R/plot.R")
plot_bayes(score_result_le$bayes_score, ordIndex_le)
```


```{r rdata Leng}
# finds the start of the cell-cycle for time-series scoring
source("reCAT_master/R/get_start.R")
cycle_start_le = get_start(score_result_le, ordIndex_le)
print(cycle_start_le)

# find best region for assessing each phase
source("reCAT_master/R/get_rdata.R")
rdata_le <- get_rdata(score_result_le, ordIndex_le)
print(rdata_le)

# obtain the real time cycle
source("reCAT_master/R/get_myord.R")
myord_le = get_myord(cycle_start_le, ordIndex_le)
```

Finds the cell cycle phases.
```{r classification Leng}
source("reCAT_master/R/get_hmm.R")
cls_num <- 3
hmm_result_le <- get_hmm_order(bayes_score = score_result_le$bayes_score, mean_score = score_result_le$mean_score, ordIndex = ordIndex_le, cls_num = cls_num, myord = myord_le, rdata = rdata_le)

g1_c_le <- sum(hmm_result_le == 1)
s_c_le <- sum(hmm_result_le == 2)
g2m_c_le <- sum(hmm_result_le == 3)

# saving the results as vectors
recat_le_pre <- c(g1_c,g2m_c,s_c)

# translating from numbers into letters
data_comp <- data.frame(hmm_result_le[order(ordIndex_le)], rownames(exprs_le_log))
le_rec_vec <- gsub("1", "G1", data_comp[[1]])
le_rec_vec <- gsub("2", "S", le_rec_vec)
le_rec_vec <- gsub("3", "G2M", le_rec_vec)

table(le_rec_vec)
```

## Sasagawa *et al* data (2013)
Section for running the data from Sasagawa.

### Load the Sasagawa data
The data is a small dataset of 23 mice cells. The ES files with a clear cell annotation is used (G1, S and M). The annotation is 8 cells in G1, 7 in S and 8 in G2M.

The data can be downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE42268. This is 23 txt files. In this code they are all put into a folder called "sasagawa" where all the txt files are loaded from.

Note: In this section you have to change the path manually.

```{r data Sasagawa}
path_sa <- paste(path, "/embryonic_data/sasagawa/", sep = "")

# reading in the separate text files and putting them next to each other in a data frame
txt_files_ls = list.files(path=path_sa, pattern="*.txt")
txt_files_ls <- paste(path_sa,txt_files_ls, sep ="")
txt_files_df <- lapply(txt_files_ls, function(x) {read.table(file = x, header = T, sep ="")})
exprs_sa <- do.call("cbind", lapply(txt_files_df, as.data.frame))

# saving the ids
id <- as.data.frame(exprs_sa[,1:2])

# removing the rows with ids from the matrix
exprs_sa <- exprs_sa[ ,-which(names(exprs_sa) %in% c("id", "gene.symbol"))]
```

### Filtering out cells of low quality
See report for details.
```{r filter Sasagawa}
# filter out cells with low nDet
nDet <- colSums(exprs_sa>1)
hist(nDet,n=100)
keep <- which(nDet>4000)

exprs_sa <- exprs_sa[,keep]

# adding ensembl ids and hgnc symbols to matrix
rownames(exprs_sa) <- paste(id[,1]) 
exprs_sa <- cbind(hgnc_symbol=id[,2],exprs_sa)
```

### Running Cyclone
Cyclone requires a matrix with ensembl IDs as input for mice data.
```{r cyclone Sasagawa}
# editing the matrix to have only ensembl ids
exprs_sa_id <- data.matrix(exprs_sa[,-1])
mmu.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))

# running cyclone
set.seed(123)
sce <- SingleCellExperiment(assays=list(counts = exprs_sa_id, logcounts = log2(exprs_sa_id+1)))
cc <- cyclone(sce, pairs=mmu.pairs)

table(cc$phase)

# saving the result as a numeric vector
cyc_sa_pre <- as.vector(unlist(lapply(c(table(cc$phases)), as.numeric)))
sa_cyc_vec <- unlist(cc$phases)
```

### Running Seurat
Seurat requires ensembl IDs as rownames as input for mice data.
```{r Seurat Sasagawa}
# create a seurat object
sdata <- CreateSeuratObject(raw.data = exprs_sa_id, min.cells = 3, min.genes = 200,
    project = "CC",is.expr=1,meta.data=NULL)

sdata <- NormalizeData(sdata)
sdata <- ScaleData(sdata)
sdata <- FindVariableGenes(object = sdata, mean.function = ExpMean, 
                           dispersion.function = LogVMR, x.low.cutoff = 0.2, 
                           x.high.cutoff = 10, y.cutoff = 0.5)

# run PCA
sdata <- RunPCA(object = sdata, pc.genes = sdata@var.genes, do.print = FALSE)  
# predict cell cycle
sdata <- CellCycleScoring(sdata, g2m.genes = cc.genes.mouse$g2m.genes, 
                          s.genes = cc.genes.mouse$s.genes)

table(sdata@meta.data$Phase)

# saving the result as a numeric vector
seu_sa_pre <- as.vector(unlist(lapply(table(sdata@meta.data$Phase), as.numeric)))
sa_seu_vec <- unlist(sdata@meta.data$Phase)
```

### Running reCAT
reCAT requires a matrix with ensembl IDs as rownames as input for mice data.

Clears out all genes not present as marker genes
```{r preprocess Sasagawa}
# log transforms the data as get_test requires log-transformed data 
exprs_sa_log <- log(exprs_sa_id+1)

source("reCAT_master/R/get_test_exp.R")
exprs_sa_log <- get_test_exp(exprs_sa_log)
```

Orders the data using the TSP algorithm
```{r order Sasagawa}
source("reCAT_master/R/get_ordIndex.R")
# exprs_sa = input matrix, 3 = no. threads used for processing, can be increased on cluster
nthread =3

ordIndex <- get_ordIndex(exprs_sa_log, nthread)
```

Get cell-cycle scores using the Bayes-scores & Mean-scores algorithm
```{r scores Sasagawa}
source("reCAT_master/R/get_score.R")
score_result <- get_score(t(exprs_sa_log))

# plots the score values 
source("reCAT_master/R/plot.R")
plot_bayes(score_result$bayes_score, ordIndex)
plot_mean(score_result$mean_score, ordIndex)
```


```{r rdata Sasagawa}
# finds the start of the cell-cycle for time-series scoring
source("reCAT_master/R/get_start.R")
cycle_start = get_start(score_result, ordIndex)
print(cycle_start)

# find best region for assessing each phase
source("reCAT_master/R/get_rdata.R")
rdata <- get_rdata(score_result, ordIndex)
print(rdata)

# obtain the real time cycle
source("reCAT_master/R/get_myord.R")
myord = get_myord(cycle_start, ordIndex)
```

Finds the cell cycle phases
```{r classification Sasagawa}
source("reCAT_master/R/get_hmm.R")
cls_num <- 3
hmm_result <- get_hmm_order(bayes_score = score_result$bayes_score, mean_score = score_result$mean_score, ordIndex = ordIndex, cls_num = cls_num, myord = myord, rdata = rdata)

g1_c = sum(hmm_result == 1)
s_c = sum(hmm_result == 2)
g2m_c = sum(hmm_result == 3)

# saves the result into a vector
recat_sa_pre <- c(g1_c, g2m_c, s_c)

#translating from numbers into letters
data_comp <- data.frame(hmm_result[order(ordIndex)], rownames(exprs_sa_log))
sa_rec_vec <- gsub("1", "G1", data_comp[[1]])
sa_rec_vec <- gsub("2", "S", sa_rec_vec)
sa_rec_vec <- gsub("3", "G2M", sa_rec_vec)

table(sa_rec_vec)
```

## Evaluation
This section is for evaluating the result of the programs. In order to run this section all programs have to have been run using all the datasets.

### RMSE
This section makes plots for each of the programs of the NRMSE values. It is calculated with the result vectors and the annotation of the cells.

```{r annotation data}
# creating vectors of the annotations
# Buettner et al. data
bu_an_vec <- unlist(M$cell_cycle_stage)
bu_an <- as.vector(unlist(lapply(table(M$cell_cycle_stage), as.numeric)))

# Leng et al. data
col <- colnames(exprs_le[,-1])
le_an_vec <- gsub('\\_.*','',col)
le_an_vec <- gsub("G2", "G2M",le_an_vec)
le_an <- as.vector(unlist(lapply(table(le_an_vec), as.numeric)))

# Sasagawa et al. data
sa_an_vec <- gsub(".*ESG1.*","G1",txt_files_ls)
sa_an_vec <- gsub(".*ESS.*","S",sa_an_vec)
sa_an_vec <- gsub(".*ESM.*","G2M",sa_an_vec)
sa_an <- as.vector(unlist(lapply(table(sa_an_vec), as.numeric)))

```

Calculatin the NRMSE value for the programs and all of the datasets.
```{r NRMSE}
# function for calculating NRMSE in %
NRMSE = function(m, o){
  sqrt(mean((m - o)^2))/sum(m)*100
}

# calculating NRMSE for cyclone
NRMSE_cyc <- c(NRMSE(cyc_bu_pre, bu_an), NRMSE(cyc_le_pre, le_an), NRMSE(cyc_sa_pre, sa_an))
 
# calculating NRMSE for Seurat
NRMSE_seu <- c(NRMSE(seu_bu_pre, bu_an), NRMSE(seu_le_pre, le_an), NRMSE(seu_sa_pre, sa_an))

# caluculating NRMSE for reCAT
NRMSE_recat <- c(NRMSE(recat_bu_pre, bu_an), NRMSE(recat_le_pre, le_an), NRMSE(recat_sa_pre, sa_an))

```

Creating barplots of the NRMSE values. The number of cells are after filtration. The axis are fixed to enable visual comparison of the NRMSE values between the programs but this can easily be changed by removing or changing ylim.
```{r barplot}
barplot(NRMSE_seu, main = "Seurat", xlab ="Data", ylab="NRMSE(%)", ylim=c(0,35), names.arg=c("Buettner et al. (272 cells)","Leng et al. (247 cells)", "Sasagawa et al. (23 cells)"), border="blue", density=c(10), col=c("purple","purple","purple"))


barplot(NRMSE_cyc, main = "Cyclone", xlab ="Data", ylab="NRMSE(%)", ylim=c(0,35), names.arg=c("Buettner et al. (272 cells)","Leng et al. (247 cells)", "Sasagawa et al. (23 cells)"), border="blue", density=c(10), col=c("purple","purple","purple"))


barplot(NRMSE_recat, main = "reCAT", xlab ="Data", ylab="NRMSE(%)", ylim=c(0,35), names.arg=c("Buettner et al. (272 cells)","Leng et al. (247 cells)", "Sasagawa et al. (23 cells)"), border="blue", density=c(10), col=c("purple","purple","purple"))
```
### Evaluation of each program
The section compares the programs with itself. The mismatch chunk produces lists where the annotation is in one column and the prediction is in the other for the cells that have been misclassified. Cells with the correct classification is removed. The rownumber represents the number of the column in the original data matrix for the cell that have been misclassified. A confusion matrix is done for every program and every data set.

## Cyclone
Section testing Cyclone on all three datasets.
```{r mismatch cyclone}
# Buettner et al. data

# comparing output to annotation
mask <- bu_an_vec == bu_cyc_vec

# creating the list and removing the correctly classified cells
bu_cyc_mat <- as.data.frame(bu_an_vec)
bu_cyc_mat <- cbind(bu_cyc_mat,bu_cyc_vec)
colnames(bu_cyc_mat) <- paste(c("actual","predicted cyclone"))

mismatch_bu_cyc <- bu_cyc_mat[!mask,]

# Leng et al. data
mask <- le_an_vec == le_cyc_vec

le_cyc_mat <- as.data.frame(le_an_vec)
le_cyc_mat <- cbind(le_cyc_mat,le_cyc_vec)
colnames(le_cyc_mat) <- paste(c("actual","predicted cyclone"))

mismatch_cyc_le <- le_cyc_mat[!mask,]

# Sasagawa et al. data
mask <- sa_an_vec == sa_cyc_vec

sa_cyc_mat <- as.data.frame(sa_an_vec)
sa_cyc_mat <- cbind(sa_cyc_mat,sa_cyc_vec)
colnames(sa_cyc_mat) <- paste(c("actual","predicted cyclone")) 

mismatch_cyc_sa <- sa_cyc_mat[!mask,]
```

Creating confusion matrices
```{r confusion matrix cyclone}
#install.packages("caret")
library(caret)

# Buettner et al. data
confusionMatrix(factor(bu_cyc_vec), factor(bu_an_vec))

# Leng et al. data
confusionMatrix(factor(le_cyc_vec),factor(le_an_vec))

# Sasagawa et al. data
confusionMatrix(factor(sa_cyc_vec),factor(sa_an_vec))
```

## Seurat
Section for evaluating Seurat.
```{r mismatch seurat}
# Buettner et al. data

# comparing output to annotation
mask <- bu_an_vec == bu_seu_vec

# creating the list and removing the correctly classified cells
bu_seu_mat <- as.data.frame(bu_seu_vec)
bu_seu_mat <- cbind(bu_an_vec,bu_seu_mat)
colnames(bu_seu_mat) <- paste(c("actual","predicted seurat"))

mismatch_bu_seu <- bu_seu_mat[!mask,]

# Leng et al. data
mask <- le_an_vec == le_seu_vec

le_seu_mat <- as.data.frame(le_seu_vec)
le_seu_mat <- cbind(le_an_vec,le_seu_mat)
colnames(le_seu_mat) <- paste(c("actual","predicted seurat"))

mismatch_le_seu <- le_seu_mat[!mask,]

# Sasagawa et al. data
mask <- sa_an_vec == sa_seu_vec

sa_seu_mat <- as.data.frame(sa_seu_vec)
sa_seu_mat <- cbind(sa_an_vec,sa_seu_mat)
colnames(sa_seu_mat) <- paste(c("actual","predicted seurat"))

mismatch_sa_seu <- sa_seu_mat[!mask,]
```

Creating confusion matrices.
```{r confusion matrix seurat}
# Buettner et al. data
confusionMatrix(factor(bu_seu_vec),factor(bu_an_vec))

# Leng et al. data
confusionMatrix(factor(le_seu_vec),factor(le_an_vec))

# Sasagawa et al. data
confusionMatrix(factor(sa_seu_vec),factor(sa_an_vec))
```

## reCAT
Section for evaluating reCAT.
```{r mismatch reCAT}
# Buettner et al. data

# comparing output to annotation
mask <- bu_an_vec == bu_rec_vec

# creating the list and removing the correctly classified cells
bu_rec_mat <- as.data.frame(bu_rec_vec)
bu_rec_mat <- cbind(bu_an_vec, bu_rec_mat)
colnames(bu_rec_mat) <- paste(c("actual","predicted reCAT"))

mismatch_bu_rec <- bu_rec_mat[!mask,]

# Leng et al. data
mask <- le_an_vec == le_rec_vec

le_rec_mat <- as.data.frame(le_rec_vec)
le_rec_mat <- cbind(le_an_vec, le_rec_mat)
colnames(bu_rec_mat) <- paste(c("actual","predicted reCAT"))

mismatch_le_rec <- le_rec_mat[!mask,]

# Sasagawa et al. data
mask <- sa_an_vec == sa_rec_vec

sa_rec_mat <- as.data.frame(sa_rec_vec)
sa_rec_mat <- cbind(sa_an_vec, sa_rec_mat)
colnames(sa_rec_mat) <- paste(c("actual","predicted reCAT"))

mismatch_sa_rec <- sa_rec_mat[!mask,]
```

Creating confusion matrices.
```{r confusion matrix reCAT}
# Buettner et al. data
confusionMatrix(factor(bu_rec_vec),factor(bu_an_vec))

# Leng et al. data
confusionMatrix(factor(le_rec_vec),factor(le_an_vec))

# Sasagawa et al. data
confusionMatrix(factor(sa_rec_vec),factor(sa_an_vec))
```