---
title: "reCAT_run"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load some packages

```{r load, warning=FALSE}
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(doParallel))
suppressPackageStartupMessages(library(mclust))
suppressPackageStartupMessages(library(cluster))
suppressPackageStartupMessages(library(TSP))
```

Load the data 

```{r data}
# load("reCAT_master/data/ola_mES_2i.RData")
cc.data <- readRDS("embryonic_data/EMTAB2805.rds")

# get tpms at gene level
exprs <- assays(experiments(cc.data)[["gene"]])[["TPM"]]

# remove spike-ins from expression table
ercc <- grep("ERCC",rownames(exprs))
exprs <- exprs[-ercc,]

# metadata with cell cycle stage assignment
M <- as.data.frame(colData(cc.data))
head(M)
```

```{r preprocess}
source("reCAT_master/R/get_test_exp.R")
test_exp <- get_test_exp(test_exp)
exprs <- assays(experiments(cc.data)[["gene"]])[["TPM"]]

# filter out cells with low nDet
# Colsums = Form row and column sums and means for numeric arrays (or data frames).
nDet <- colSums(exprs>1)
hist(nDet,n=100)
# make a cutoff at 4000 genes, removes 16 cells
keep <- which(nDet>4000)

exprs <-exprs[,keep]
```

Get order of cycle

```{r order}
source("reCAT_master/R/get_ordIndex.R")
ordIndex <- get_ordIndex(test_exp, 1)
```

Get cell-cycle scores

```{r scores}
source("reCAT_master/R/get_score.R")
score_result <- get_score(t(test_exp))
score_result$bayes_score
score_result$mean_score
```

Order cells based on cycle usign HMM 

```{r hmm}
source("reCAT_master/R/get_hmm.R")
load("embryonic_data/EMTAB2805.rds")
myord = c(4:1, 295:5)
hmm_result <- get_hmm_order(bayes_score = score_result$bayes_score, 
	mean_score = score_result$mean_score, 
	ordIndex = ordIndex, cls_num = 3, myord = myord, rdata = rdata)
```


# get tpms at gene level

# The Assays virtual class and its methods provide a formal abstraction of the assays slot of SummarizedExperiment object
# The RangedSummarizedExperiment class is a matrix-like container where rows represent ranges of interest and columns represent samples (with sample data summarized as a DataFrame).
# A RangedSummarizedExperiment contains one or more assays, each represented by a matrix-like object of numeric or other mode.
# exprs <- assays(experiments(cc.data)[["gene"]])[["TPM"]]
# remove the .1 etc from the ensembl gene name
# gname <- rownames(exprs)
# ensname <- gsub("\\.\\d+","",gname)
# rownames(exprs) <- ensname

# remove spike-ins from expression table
# ercc <- grep("ERCC",rownames(exprs))
# exprs <- exprs[-ercc,]
# exprs[is.na(exprs)] <- 0

# metadata with cell cycle stage assignment
# colData: Each column of a SummarizedExperiment object represents a sample. Information about the samples are stored in a DataFrame, accessible using the function colData
# M <- as.data.frame(colData(cc.data))
# head(M)