---
title: "Detecting cell-cycle stage difference in glioblastoma cells"
author: "Donghyung Lee"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{"Detecting cell-cycle stage difference in glioblastoma cells"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Here, we illustrate how to use the iasva package to detect cell cycle stage
difference within single cell RNA sequencing data. We use single cell RNA
sequencing (scRNA-Seq) data obtained from human glioblastoma samples
([Petel et. al., 2014](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4123637/)).
This dataset is included in a R data package ("iasvaExamples") containing data
examples for IA-SVA (https://github.com/dleelab/iasvaExamples). 
To install the package, follow the instruction provided in the GitHub page.

## Install packages
```{r install_packages, echo=TRUE, eval=FALSE}
#devtools
library(devtools)
#iasva
devtools::install_github("UcarLab/iasva")
#iasvaExamples  
devtools::install_github("dleelab/iasvaExamples")
```

## Load packages
```{r load_packages, echo=TRUE, message=FALSE}
rm(list=ls())
library(irlba) # partial SVD, the augmented implicitly restarted Lanczos bidiagonalization algorithm
library(iasva)
library(iasvaExamples)
library(sva)
library(SCnorm)
library(scran)
library(scater)
library(Rtsne)
library(pheatmap)
library(corrplot)
library(DescTools) #pcc i.e., Pearson's contingency coefficient
library(RColorBrewer)
library(SummarizedExperiment)
library(vioplot)
color.vec <- brewer.pal(3, "Set1")

# Normalization.
normalize <- function(counts) 
{
    normfactor <- colSums(counts)
    return(t(t(counts)/normfactor)*median(normfactor))
}
```

## Load the Buettner single cell RNA-Seq data
```{r load_data, echo=TRUE}
buettner.data <- readRDS("/Users/erikh/Desktop/git/scrna-cell-cycle/embryonic_data/EMTAB2805.rds")

# get tpms at gene level
buet_exprs <- assays(experiments(buettner.data)[["gene"]])[["TPM"]]
dim(buet_exprs)
# remove the .1 etc from the ensembl gene name
gname <- rownames(buet_exprs)
ensname <- gsub("\\.\\d+","",gname)
rownames(buet_exprs) <- ensname

# remove spike-ins from expression table
ercc <- grep("ERCC",rownames(buet_exprs))
buet_exprs <- buet_exprs[-ercc,]

# metadata with cell cycle stage assignment
M <- as.data.frame(colData(buettner.data))
head(M)
dim(buet_exprs)
```

## Filter out clearly low quality cells from Buettner data

Clearly some outlier cells on negative PC1 with low number of genes with pca using full dataset.
Mainly S-phase cells, even if most S-phase cells have a high median expression.

```{r filter}
# filter out cells with low nDet
nDet <- colSums(buet_exprs>1)
hist(nDet,n=100)
# make a cutoff at 4000 genes, removes 16 cells
keep <- which(nDet>4000)

buet_exprs<-buet_exprs[,keep]
M <- M[keep,]
dim(buet_exprs)

```

## Normalization for Buettner data
```{r normalization, fig.width=8, fig.height=6}
## count-depth relationship for all genes
norm_buet <- SingleCellExperiment(assays=list(counts = buet_exprs, logcounts = log2(buet_exprs+1)))
dim(norm_buet)
```

## Calculate the number of detected genes 
It is well known that the number of detected genes in each cell explains
a very large portion of variability in scRNA-Seq data 
([Hicks et. al. 2015 BioRxiv](http://biorxiv.org/content/early/2015/08/25/025528),
[McDavid et. al. 2016 Nature Biotechnology](http://www.nature.com/nbt/journal/v34/n6/full/nbt.3498.html)).
Frequently, the first principal component of log-transformed scRNA-Seq read counts is highly correlated
with the number of detected genes (e.g., r > 0.9). Here, we calculate the number of detected genes 
for glioblastoma cells, which will be used as an known factor in the IA-SVA analyses. 
```{r num_detected_genes, echo=TRUE, fig.width=7, fig.height=4}
Num_Detected_Genes_buet <- colSums(buet_exprs>0)
Geo_Lib_buet <- colSums(log(buet_exprs+1))
summary(Geo_Lib_buet)
barplot(Geo_Lib_buet, xlab="Cell", las=2, ylab = "Geometric library size")
lcounts_buet <- log(buet_exprs + 1)
# PC1 and Geometric library size correlation
pc1_buet = irlba(lcounts_buet - rowMeans(lcounts_buet), 1)$v[,1] ## partial SVD
cor(Num_Detected_Genes_buet, pc1_buet)
cor(Geo_Lib_buet, pc1_buet)
```

## Run IA-SVA
Here, we run IA-SVA using Geo_Lib_Size as a known factor and identify five hidden factors. 
SVs are plotted in a pairwise fashion to uncover which SVs can seperate cells. 
```{r run_iasva, echo=TRUE, fig.width= 7, fig.height=6}
set.seed(3445)
mod_buet <- model.matrix(~Geo_Lib_buet)
summ_exp_buet <- SummarizedExperiment(assays = buet_exprs)
iasva.res_buet <- iasva(summ_exp_buet, as.matrix(mod_buet[,-1]),verbose=FALSE, permute=FALSE, num.sv=7)
iasva.sv_buet <- iasva.res_buet$sv

Cluster_buet <- as.factor(iasva.sv_buet[,7] < 0.1) 
levels(Cluster_buet)=c("Cluster1","Cluster2")
table(Cluster_buet)

pairs(iasva.sv_buet[,1:7], main="IA-SVA", pch=21, col=color.vec[Cluster_buet],
      bg=color.vec[Cluster_buet], oma=c(4,4,6,14))
legend("right", levels(Cluster_buet), fill=color.vec, bty="n")

plot(iasva.sv_buet[,c(7,1)], main="IA-SVA", pch=21, xlab="SV1", ylab="SV7",
     col=color.vec[Cluster_buet], bg=color.vec[Cluster_buet])

cor(Num_Detected_Genes_buet, iasva.sv_buet[,1])
cor(Geo_Lib_buet, iasva.sv_buet[,1])
corrplot(cor(iasva.sv_buet))
```

## Find marker genes for the detected heterogeneity (SV1).
Here, using the find_markers() function we find marker genes that are significantly
associated with SV1 (multiple testing adjusted p-value < 0.05, 
default significance cutoff, and R-squared value > 0.3).  
```{r find_markers_SV1, echo=TRUE, fig.width=6, fig.height=16}
# try different R2 thresholds
pdf(paste0("/Users/erikh/Desktop/git/scrna-cell-cycle/r_scripts/erik/ia_sva/sc_output/Clustering_analyses_figure3_sv7_buet.pdf"))
r2.results_buet <- study_R2(summ_exp_buet, iasva.sv_buet, selected.svs=2, no.clusters=2)
dev.off()

marker.counts.SV7_buet <- find_markers(summ_exp_buet, 
                            as.matrix(iasva.sv_buet[,c(7)]), rsq.cutoff = 0.4)
marker.counts.SV7.long_buet <- find_markers(summ_exp_buet, 
                              as.matrix(iasva.sv_buet[,c(7)]), rsq.cutoff = 0.3)
nrow(marker.counts.SV7_buet) 
nrow(marker.counts.SV7.long_buet)

anno.col2_buet <- data.frame(Cluster=Cluster_buet, SV7=iasva.sv_buet[,7])
rownames(anno.col2_buet) <- colnames(marker.counts.SV7_buet)
head(anno.col2_buet)

cluster.col_buet <- color.vec[1:2]
names(cluster.col_buet) <- as.vector(levels(Cluster_buet))
anno.colors_buet <- list(Cluster=cluster.col_buet)
anno.colors_buet

pheatmap(log(marker.counts.SV7_buet+1), show_colnames =FALSE, 
         clustering_method = "ward.D2",cutree_cols = 2,annotation_col = anno.col2_buet,
         annotation_colors = anno.colors_buet)

pheatmap(log(marker.counts.SV7.long_buet+1), show_colnames =FALSE, 
         clustering_method = "ward.D2",cutree_cols = 2,annotation_col = anno.col2_buet,
         annotation_colors = anno.colors_buet)

gene.list_buet <- rownames(marker.counts.SV7_buet)
write.table(gene.list_buet, file = paste0("/Users/erikh/Desktop/git/scrna-cell-cycle/r_scripts/erik/ia_sva/sc_output/CC_genes.short_buet.txt"),
            col.names =F, row.names = F, quote = F)

gene.list.long_buet <- rownames(marker.counts.SV7.long_buet)
write.table(gene.list.long_buet, file = paste0("/Users/erikh/Desktop/git/scrna-cell-cycle/r_scripts/erik/ia_sva/sc_output/CC_genes.long_buet.txt"),
            col.names =F, row.names = F, quote = F)

```

## Cell type assignment using scran R package
```{r run_scran, echo=TRUE, fig.width=6, fig.height=4}
mmu.pairs <- readRDS(system.file("exdata", 
                                "mouse_cycle_markers.rds", package="scran"))
assigned_buet <- cyclone(norm_buet, pairs=mmu.pairs)
head(assigned_buet$scores)
table(assigned_buet$phases)
phase_buet <- rep("S", ncol(norm_buet))
phase_buet[assigned_buet$scores$G1 > 0.5 & assigned_buet$scores$G2M < 0.5] <- "G1"
phase_buet[assigned_buet$scores$G1 < 0.5 & assigned_buet$scores$G2M > 0.5] <- "G2M"
phase_buet[assigned_buet$scores$G1 < 0.5 & assigned_buet$scores$G2M < 0.5] <- "S"
phase_buet[assigned_buet$scores$G1 > 0.5 & assigned_buet$scores$G2M > 0.5] <- "Unknown"
table(phase_buet)

G1_buet_SV <- matrix(iasva.sv_buet[,c(1:7)][phase_buet=="G1"], ncol=7)
S_buet_SV <- matrix(iasva.sv_buet[,c(1:7)][phase_buet=="S"], ncol=7)
G2M_buet_SV <- matrix(iasva.sv_buet[,c(1:7)][phase_buet=="G2M"], ncol=7)
Unknown_buet_SV <- matrix(iasva.sv_buet[,c(1:7)][phase_buet=="Unknown"], ncol=7)

corrplot(cor(G1_buet_SV))
corrplot(cor(S_buet_SV))
corrplot(cor(G2M_buet_SV))

vioplot(G1_buet_SV[,c(1)], S_buet_SV[,c(1)], G2M_buet_SV[,c(1)], names=c("G1", "S", "G2M"), col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor (SV1)")
vioplot(G1_buet_SV[,c(2)], S_buet_SV[,c(2)], G2M_buet_SV[,c(2)], names=c("G1", "S", "G2M"), 
   col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor (SV2)")
vioplot(G1_buet_SV[,c(3)], S_buet_SV[,c(3)], G2M_buet_SV[,c(3)], names=c("G1", "S", "G2M"), 
   col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor (SV3)")
vioplot(G1_buet_SV[,c(4)], S_buet_SV[,c(4)], G2M_buet_SV[,c(4)], names=c("G1", "S", "G2M"), 
   col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor (SV4)")
vioplot(G1_buet_SV[,c(5)], S_buet_SV[,c(5)], G2M_buet_SV[,c(5)], names=c("G1", "S", "G2M"), 
   col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor (SV5)")
vioplot(G1_buet_SV[,c(6)], S_buet_SV[,c(6)], G2M_buet_SV[,c(6)], names=c("G1", "S", "G2M"), 
   col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor (SV6)")
vioplot(G1_buet_SV[,c(7)], S_buet_SV[,c(7)], G2M_buet_SV[,c(7)], names=c("G1", "S", "G2M"), 
   col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor (SV7)")
```

## Correlation between SV1 and the geometric library size
```{r SV1_geometric_lib_size, echo=TRUE}
cor(Num_Detected_Genes_buet, iasva.sv_buet[,c(1:7)])
cor(Geo_Lib_buet, iasva.sv_buet[,c(1:7)])
```
By allowing correlation between factors, IA-SVA accurately detects the cell
cycle stage difference, which is moderately correlated (|r|=0.44) with the geometric
library size (the first principal component). Existing methods fail to detect
the heterogeneity due to the orthogonality assumption.

```{r  gen_Figure3ABCDE, fig.width=5, fig.height=8}
pdf(file=paste0("/Users/erikh/Desktop/git/scrna-cell-cycle/r_scripts/erik/sc_output/Patel_Glioblastoma_MGH30_CellCycle_Figure3ABCD_buet.pdf"), width=5, height=8)
layout(matrix(c(1,2,3,4,5,5), nrow=3, ncol=2, byrow=TRUE))
plot(iasva.sv_buet[,c(1,7)], main="IA-SVA", pch=21, xlab="SV1", ylab="SV7",
     col=color.vec[Cluster_buet], bg=color.vec[Cluster], oma=c(4,4,6,12))
legend("topright", levels(Cluster_buet), border="white", fill=color.vec, bty="n")
plot(pca.res[,1:2], main="PCA", pch=21, xlab="PC1", 
     ylab="PC2", col=color.vec[Cluster], bg=color.vec[Cluster])
plot(sva.res[,1:2], main="USVA", xlab="SV1", ylab="SV2",
     pch=21, col=color.vec[Cluster], bg=color.vec[Cluster])
plot(tsne.res$Y, main="tSNE", xlab="Dimension 1", 
     ylab="Dimension 2", pch=21, col=color.vec[Cluster], bg=color.vec[Cluster])
vioplot(G1, S, G2M, Unknown, names=c("G1", "S", "G2M", "Unknown"), 
   col="gold")
title(xlab="Cell-cycle stage predictions", ylab="IA-SVA factor")
dev.off()
```